Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
modified postcondition of traverse
rewrote the deletion problem

\subsection{Templates vs. Internal Reorganization}
\label{internal_reorganize}
The \lstinline{delete} operation on binary search trees presents difficulties in applying the template approach. The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert} or \lstinline{lookup}). When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown_left} operation, %include figure?
rotating three-node sections of the tree until the node to be deleted is at a leaf. This operation is simultaneously concurrency-aware (we must lock the nodes involved to avoid race conditions) and data-structure-specific (we need to know precisely which nodes to target and how to rearrange them). This contrasts with linked-list deletion, which only involves the target node and its immediate neighbors. In practice, Krishna et al.'s proofs for tree structures~\cite{templates} either do not include deletion at all, or use a very lightweight form of deletion in which nodes are marked as ``deleted'' but are never actually removed from the data structure. \wm{compare what they say and do for split in a B+-tree}

We believe this represents a fundamental limitation in the template approach. Operations in the style of \lstinline{pushdown_left} are ubiquitous in tree structures (e.g., the rebalancing operations of AVL and red-black trees), and always weave together synchronization and restructuring in a way that cannot easily be decomposed. This is a serious impediment to the goal of verifying concurrency control separately from the data structure: even if we have a verified template and a verified single-threaded tree, we still need to verify a complex data-structure-specific concurrent function. %in fact, usually the most complicated part!
%what's the conclusion here? can we see a way through?

>>>>

\subsection{Templates vs. Internal Reorganization}
\label{internal_reorganize}
The \lstinline{delete} operation on binary search trees presents difficulties in applying the template approach. The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert} or \lstinline{lookup}). Likewise, in the data structures verified by Krishna et al., deletion can be performed by traversing to the node to be deleted, acquiring its lock and possibly its parent's lock, and then performing a local, sequential data-structure-specific operation. When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown_left} operation, %include figure or code?
rotating three-node sections of the tree until the node to be deleted is at a leaf. This operation is simultaneously concurrency-aware (we must lock the nodes involved to avoid race conditions) and data-structure-specific (we need to know precisely which nodes to target and how to rearrange them).

It is instructive to compare BST deletion with Krishna et al.'s discussion of splitting in B-link trees. They assume the existence of a maintenance thread that constantly crawls the tree searching for full nodes and splitting them. The splitting can be defined as a sequential, data-structure-specific operation on a node and its parent; the crawl itself is a variant of the \lstinline{traverse} operation, and while it is not verified in their existing proofs, we expect that it could be defined as a template function. We might similarly imagine refactoring a lock-coupling \lstinline{pushdown_left} into a template-level traversal and a data-structure-level rotation operation. However, in the BST, rotation involves three nodes (the node to be deleted and both of its children)---and, crucially, \emph{we only know that it involves three nodes once we know the target data structure is a BST}. Thus, there is no clear way to define a template function that acquires enough locks (or otherwise performs enough synchronization) to call out to a sequential rotation operation without making the template specific to BSTs.

We believe this represents a fundamental limitation in the template approach. Operations in the style of \lstinline{pushdown_left} are ubiquitous in tree structures (e.g., the rebalancing operations of AVL and red-black trees), and always weave together synchronization and restructuring in a way that cannot easily be decomposed. This is a serious impediment to the goal of verifying concurrency control separately from the data structure: even if we have a verified template and a verified single-threaded tree, we still need to verify a complex data-structure-specific concurrent function to support deletion, which is often the most complicated part of a concurrent search structure.
%what's the conclusion here? can we see a way through?
